---
title: "Lecture 2 - Statistical Rethinking"
author: "sbsambado"
date: "4/8/2020"
output: html_document
---

Code from video lectures held by Dr. Richard McElreath


Computer the posterior
1. Analytically
2. Grid approximization
3. Quadratic approximization
4. Markov chain monte carlo
```{r}
# Grid approximization


p_grid <- seq(0, 1, length.out = 1000) # define the grid
prob_p <- rep(1,1000) # define prior probability
prob_data <- dbinom(6, size = 9, prob = p_grid) 
posterior <- prob_data * prob_p # gives the curve the shape
posterior <- posterior / sum(posterior) # standardize it

plot(prob_data)
plot(posterior)
```

R code 3.12 how to calculate intervals
```{r}
# 50% percentile interval
PI(samples, prob = 0.5)

# 50% HDPI
HDPI(samples, prob = 0.5)
```

R code 3.21 :posterior predictions (merge data from previous probability events)
```{r}
nw <- rbinom(1e4, size = 9, prob = samples)
```

Homework 2M1: Compute and plot the grid approximate posterior distribution for each of the following set of observations. In each case, assume a uniform prior for p.

(1) W,W,W
```{r}
## 2.2 Building a model

# globe tossing data in a tibble
# toss a globe, what does your finger land on water 'w' or land 'l'
e <- tibble(toss = c('w', 'w','w'))

## 2.2.2 Bayesian updating
y<- (e %>%
  mutate(n_trials = 1:3, # change trial number by toss # cumulative number of trials
         n_success = cumsum(toss == 'w'))) # cumulative number of successes

sequence_length <- 50

#View(y) to make sure you added those columns to the dataset 'e'
y %>%
  expand(nesting(n_trials, toss, n_success),
         p_water = seq(0, 1, length.out = sequence_length)) %>%
  group_by(p_water) %>%
  mutate(lagged_n_trials = lag(n_trials, k = 1),
         lagged_n_success = lag(n_success, k = 1)) %>%
  ungroup() %>%
  mutate(prior = ifelse(n_trials == 1, 0.5,
                        dbinom(x = lagged_n_success,
                               size = lagged_n_trials,
                               prob = p_water)),
         likelihood = dbinom(x = n_success,
                             size = n_trials, 
                             prob = p_water),
         strip = str_c('n = ', n_trials)) %>%
  
  # normalize the prior and the likelihood,
  # putting them both in a probability metric
  group_by(n_trials) %>%
  mutate(prior = prior / sum(prior),
         likelihood = likelihood / sum(likelihood)) %>%
  # plot
  ggplot(aes(x = p_water)) +
  geom_line(aes(y = prior), linetype = 2) +
  geom_line(aes(y = likelihood)) +
  scale_x_continuous('proportion water', breaks = c(0, 0.5, 1)) +
  scale_y_continuous('plausibility', breaks = NULL) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~strip, scales = 'free_y') 
```

(2) W,W,W,L
```{r}
f <- tibble(toss = c('w', 'w','w','l'))


y<- (f %>%
  mutate(n_trials = 1:4, # change trial number by toss #
         n_success = cumsum(toss == 'w')))

sequence_length <- 50
#View(y)
y %>%
  expand(nesting(n_trials, toss, n_success),
         p_water = seq(0, 1, length.out = sequence_length)) %>%
  group_by(p_water) %>%
  mutate(lagged_n_trials = lag(n_trials, k = 1),
         lagged_n_success = lag(n_success, k = 1)) %>%
  ungroup() %>%
  mutate(prior = ifelse(n_trials == 1, 0.5,
                        dbinom(x = lagged_n_success,
                               size = lagged_n_trials,
                               prob = p_water)),
         likelihood = dbinom(x = n_success,
                             size = n_trials, 
                             prob = p_water),
         strip = str_c('n = ', n_trials)) %>%
  group_by(n_trials) %>%
  mutate(prior = prior / sum(prior),
         likelihood = likelihood / sum(likelihood)) %>%
  
  ggplot(aes(x = p_water)) +
  geom_line(aes(y = prior), linetype = 2) +
  geom_line(aes(y = likelihood)) +
  scale_x_continuous('proportion water', breaks = c(0, 0.5, 1)) +
  scale_y_continuous('plausibility', breaks = NULL) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~strip, scales = 'free_y')
```

(3) L, W, W, L, W, W, W
```{r}
g <- tibble(toss = c('l','w', 'w','l','w','w','w'))

y<- (g %>%
  mutate(n_trials = 1:7, # change trial number by toss #
         n_success = cumsum(toss == 'w')))

sequence_length <- 50
#View(y)
y %>%
  expand(nesting(n_trials, toss, n_success),
         p_water = seq(0, 1, length.out = sequence_length)) %>%
  group_by(p_water) %>%
  mutate(lagged_n_trials = lag(n_trials, k = 1),
         lagged_n_success = lag(n_success, k = 1)) %>%
  ungroup() %>%
  mutate(prior = ifelse(n_trials == 1, 0.5,
                        dbinom(x = lagged_n_success,
                               size = lagged_n_trials,
                               prob = p_water)),
         likelihood = dbinom(x = n_success,
                             size = n_trials, 
                             prob = p_water),
         strip = str_c('n = ', n_trials)) %>%
  group_by(n_trials) %>%
  mutate(prior = prior / sum(prior),
         likelihood = likelihood / sum(likelihood)) %>%
  
  ggplot(aes(x = p_water)) +
  geom_line(aes(y = prior), linetype = 2) +
  geom_line(aes(y = likelihood)) +
  scale_x_continuous('proportion water', breaks = c(0, 0.5, 1)) +
  scale_y_continuous('plausibility', breaks = NULL) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~strip, scales = 'free_y')
```

2.3 Components of the model
1. A likelihood function: the number of ways each conjecture could produce an observation
2. One or more parameters: accumulated numbers of ways each conjecture could produce the entire data
3. A prior: the initial plausibility of each conjectured cause of data

2.3.1 Likelihood
```{r}
dbinom(x = 6, size = 9, prob = 0.5)

# but let's change values of 'prob'

tibble(prob = seq(0, 1, by = 0.01)) %>%
  ggplot(aes(x = prob,
             y = dbinom(x = 6, size = 9, prob = prob))) +
  geom_line() +
  labs(x = 'probability',
       y = 'binomial likelihood') +
  theme(panel.grid = element_blank())
```

2.3.2 Parameters

Data are measures and known.
Parameters are unknown and must be estimated from data.

2.3.3 Prior