---
title: "Lec 3 - Statistical Rethinking Richard McElreath"
author: "sbsambado"
date: "4/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Linear regression
- model of mean and variance of normal (Gaussian) distribution
- mean as addative combination of weighted variance
- constant variance


  W ~ Binomal(N, p)
  p ~ Uniform(0,1)
  
  
Kalahar foragers example

Gaussian model (Normal distribution)
```{r}
library(rethinking)
data("Howell1")
d <- Howell1

precis(d)


# prior predictive distribution
# what do these priors implay about height, before we see data?

sample_mu <- rnorm(1e4, 178, 20)
sample_sigma <- runif(1e4, 0, 50)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)
```

Quadratic approximation
- approximate posterior as Gaussian
- can estimate with two things
  - peak of posterior, maximum a posteriori (MAP)
  - standard deviations and correlations of parameters (covariance matrix)
- with flat priors, same as conventional maximum likelihood estimation


Using `quap`

hi ~ Normal(mu, sd)
mu ~ Normal(178, 20)
sd ~ Uniform(0,50)
```{r}
# R code 4.27

flist <- alist(
  height ~ dnorm(mu, sigma),
  mu ~ dnorm(178, 20),
  sigma ~ dunif(0, 50)
)

# R code 4.28

m4.1 <- quap(flist, data = d) # define Gaussian distribution

precis(m4.1) # summary command


post <- extract.samples(m4.1, n = 1e4)
head(post)
```

Scaffolds
- `quad` is a scaffold
  - force full specification of model, so you'll learn it
  - works with a very wide calss of models
  - same as penalized maximum likelihood
  - not always a good way to approximate posterior
  
  
Adding a predictor variable
- how does weight describe height?
- use a linear model of the mean, mu

hi ~ Normal(mui, sd) [likelihood]
mui ~ alpha + beta(xi - x_bar) [linear model]
alpha ~ Normal(178, 20) [alpha prior]
beta ~ Normal(0,10) [beta prior]
sd ~ Uniform(0, 50) [sd prior]

```{r}

# R code 4.38
set.seed(2971)
N <- 100 # 100 lines
a <- rnorm(N, 178, 20)
b <- rnorm(N, 0, 10)


# R code 4.40
# log distributed data how to make data not negative
# log distribution is normal when you log it (so you have all positive real values)
b <- rlnorm(1e4, 0, 1)
dens(b, xlim = c(0,5), adj = 0.1)

set.seed(2971)
N <- 100 # 100 lines
a <- rnorm(N, 178, 20)
b <- rnorm(N, 0, 1)

```

Approximate the posterior
```{r}
# R code 4.42

data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18,] # all adults

# define average weight, x-bar
xbar <- mean(d2$weight)

# fit model
m4.3 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar), # log posterior
    a ~ dnorm(178, 20),
    b ~ dlnorm(0,1),
    sigma ~ dunif(0, 50)
  ),
  data = d2)
# R.code 4.44
precis(m4.3)

plot(height ~ weight, data = d2, col = rangi2)
post <- extract.samples(m4.3)
a_map <- mean(post$a)
b_map <- mean(post$b)
curve(a_map + b_map*(x - xbar), add = TRUE)
```

