---
title: "Statistical_Rethinking_Ch1,2"
author: "sbsambado"
date: "3/30/2020"
output: html_document
---
Install rethink R package
```{r}
install.packages(c('coda', 'mvtnorm', 'devtools', 'dagitty'))
library(devtools)

devtools :: install_github('rmcelreath/rethinking')
install.packages('tidyverse')
library(tidyverse)

library(git2r)
```

Install rstan 
1. Update to macOS Catalina (10.15.4)
2. Download macos-rtools-3.2.2pkg (this has Xcode built in) from github
3. To install, use cntrl + click and select 'Open' from menu (do not double-click)
4. Install rstan
```{r}
## if you followd Installing Rstan instruction (https://github.com/stan-dev/rstan/wiki/Installing-RStan-from-source-on-a-Mac#prerequisite--c-toolchain-and-configuration)

# prerequisite: C++ toolchain and configuration
install.packages('Rcpp', repos = 'https://rcppcore.github.io/drat')

install.packages('Rcpp', dependencies = TRUE) # saw in another chain if you add 'depdencies' argument it overrides some bullshit

# remove the package, which I don't get and I know I didn't have previously
remove.packages('rstan')
if( file.exists('.RData')) file.remove('.RData')

# det desired number of cores to use during installation
Sys.setenv(MAKEFLAGS = '-j4')

# install the CRAN version of RStan from source
install.packages('rstan', type = 'source')

# or install the development version of Rstan from Github
remotes::install_github('stan-dev/rstan', ref = 'develop',
                        subdir = 'rstan/rstan', build_opts = '')
# ^^ these commands didn't work for me

# troubleshooted online and saw this posted by bgoodri (creator of RStan)
install.packages('rstan', repos = 'https://cran.r-project.org', dependencies = TRUE)
# ^ that seemed to work and was able to call library(rstan)
library(rstan)
```
^^ I think I have successfully installed Rstan and rethinkingstatistics github, honestly that was a cluster fuck

Now my github connection was lost, need to reinstall and connect to my github repo
```{r}
install.packages('git2r')
library(git2r)
??'git2r-package'

install_github()

## will come back to this later

```

Test rethinking
```{r}

library(rstan)
options(mc.cores = parallel ::detectCores())
rstan_options(auto_write = TRUE)

library(rethinking)

# testing rethinking

data(rugged)
d <- rugged
d$log_gdp <- log(d$rgdppc_2000)
dd <- d[ complete.cases(d$rgdppc_2000) , ]
dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp)
dd$rugged_std <- dd$rugged / max(dd$rugged)
dd$cid <- ifelse( dd$cont_africa==1 , 1 , 2 )

## R code 9.12
m8.3 <- quap(
  alist(
    log_gdp_std ~ dnorm( mu , sigma ) ,
    mu <- a[cid] + b[cid]*( rugged_std - 0.215 ) ,
    a[cid] ~ dnorm( 1 , 0.1 ) ,
    b[cid] ~ dnorm( 0 , 0.3 ) ,
    sigma ~ dexp( 1 )
  ) , data=dd )
precis( m8.3 , depth=2 )

## R code 9.13
dat_slim <- list(
  log_gdp_std = dd$log_gdp_std,
  rugged_std = dd$rugged_std,
  cid = as.integer( dd$cid )
)
str(dat_slim)
View(dat_slim)

## R code 9.14
## This bit calls stan, here's where you'll have trouble if your c compiler isn't working
m9.1 <- ulam(
  alist(
    log_gdp_std ~ dnorm( mu , sigma ) ,
    mu <- a[cid] + b[cid]*( rugged_std - 0.215 ) ,
    a[cid] ~ dnorm( 1 , 0.1 ) ,
    b[cid] ~ dnorm( 0 , 0.3 ) ,
    sigma ~ dexp( 1 )
  ) , data=dat_slim , chains=1 )
#Some simple output from a simple stan model
print(m9.1)
# can't find the function ulam so I think I just need to install something else
```
test stan
```{r}
schools.code <- 'data{
  int<lower=0> J;
  real y[J];
  real<lower=0> sigma[J];
}
parameters {
real mu;
real<lower=0> tau;
vector[J] eta;
}
transformed parameters {
vector[J] theta = mu + tau * eta;
}
model {
target += normal_lpdf(eta | 0, 1);
target += normal_lpdf( y | theta, sigma);
}
'

schools_dat <- list(J = 8,
                    y = c(28, 8, -3, 7, -1, 1, 18, 12),
                    sigma = c(15, 10, 16, 11, 9, 11, 10, 18))
fit <- stan(model_code = schools.code, data = schools_dat, chains = 4, iter = 10000, control = list(adapt_delta = 0.99))
print(fit)
plot(fit, pars = c('eta'))
```

